A parallel runtime framework for communication intensive stream applications

### 2013-05-10 paper.tex
final submission for ispa 2013

In this paper we propose a parallel processing framework for building stream applications that are bound by very large communication. A simple framework is used, where the filters are described as \textit{actors} and the application is represented by connecting such \textit{actors} into a directed acyclic graph (DAG). Then our framework is used to establish, (1)the communication system that allows these \textit{actors} to send and receive data across each other and (2)the computation system that executes these tasks in parallel on a given multicore system. The main contributions can be described as,

@@

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using the multiple queues makes it possible to do much of the communication in parallel and by using the multiple cores, the filters from the stream application can be executed concurrently.

A major bottleneck for communication on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research uses alternate mechanisms to access data from the network hardware directly to overcome this problem\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with specialized system software that allows direct access to hardware-controlled packet queues, dramatically reducing overheads and improving performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on a \textit{flow}, a specific \textit{stream} between two filters. Data in a \textit{stream} must be processed sequentially, whereas different \textit{flows} can often be processed in parallel.

Newer network hardware has introduced a mechanism for separating arriving data into multiple queues, where data that belong to the same \textit{flow} are\comment{ guaranteed to be} placed in the same queue\cite{micro2008} \cite{intel2010}. We can utilize this feature to operate on multiple streams of data by processing each of the queues in parallel. However, building applications that can use this feature directly is a challenging task because it requires explicit parallel programming and interaction with low-level network interfaces. Previous research~\cite{Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207} that focuses on improving the performance of network applications in such a manner shows the complexity involved in building such a system.

@@

Our framework provides support for sending and receiving data in parallel over the NIC from user space and a common parallel execution engine that supports both application and packet processing tasks.

\subsection{Parallel Packet I/O}
\label{ppio}
The parallel packet I/O interface, when combined with the application tasks, provides concurrent communication for the application threads or processes over the network. To do this, two important requirements have to be met. Firstly, access to network data from user space should be provided, which can be read from and written to in parallel. Secondly, it should be possible to integrate the packet operations with the application by providing the necessary interfaces for performing these packet operations, thereby allowing for some flexibility in the choice of communication protocol.

\subsubsection{Userspace Packet Access}
Applications executed in kernel space for accessing hardware devices directly would lead to system instability. Instead, previous researchers \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access to network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in kernel space and user space to export the network hardware buffer's memory region. This enables applications in user space to directly access the network hardware. It also prevents applications from accessing memory outside valid regions by checking the limits of the ring buffer. Packets can be written to the buffers when space is available, and an \textit{ioctl} call to the driver enables the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in user space to check when new packets are available. On reading or writing packets, the available space is updated.

@@

\subsubsection{Packet Processing Operations}

Packet processing operations consist of converting the raw data that the application provides into network packets suitable for transmission. Since the application tasks are represented as a set of operations on a data stream, we include these operations as stages through which the data flows through to get transformed into data packets. Integrating them with the application task graph makes it easier to extract parallelism later in the execution engine. Several user space libraries exist that perform packet related processing operations like libnids \cite{libnids}, lwip \cite{Dunkels01designand}, Click \cite{Kohler2000} etc. Since our application is targeted at a closed environment, we chose the Click modular router due to its flexibility and vast library of elements. Our current framework supports the UDP protocol but Click's modular infrastructure makes it easy to support more complex ones. \comment{Indeed, stream applications with more complex protocols would benefit even more from the framework due to the parallelizing of these complex tasks.}

@@

In order to prevent unwanted process migration due to the OS scheduler, we specify \textit{affinity} of the MPI processes to particular CPU cores. This prevents each process from being executed on a different CPU other than that specified. This makes the processes act as execution engines, running on each of the cores on which we schedule the operations that are to be performed. Previous research \cite{Salehi:1996:EAS:234766.234769}\cite{1409136} highlights the importance of affinity in network sensitive applications. Next, by utilizing our \textit{Parallel Packet I/O}, we specify the hardware queues each of these processes handle. \comment{Initially we divide the queues evenly among the different processes we create, but later during execution we could modify the allocation based on the application operations scheduled on them. Our modifications to \texttt{netmap}'s hardware queue assignment API can help us here in redistributing the queues to different processes.} By using MPI's \texttt{send} and \texttt{receive} inter-process communication calls to handle communication between execution engines located at each CPU, it is possible to modify the attributes of the process such as CPUs assigned or number of NIC queues handled, and even the operations the processes were initially assigned.

\subsubsection{Parallel task processing}
\label{ppt}

Once we have scheduled our execution engines on the physical system, the next step is to execute the stream graph for different \textit{flows}. To extract data parallelism, we replicate the stream graph onto multiple execution engines. This ensures operations on the same \textit{flow} are performed on the same CPU, thereby using local cache. By making sure that packets belonging to a specific f\textit{flow} always end up in the same execution engine, we only perform operations conforming to that stream of data. \comment{We use the ability of modern NICs to classify packets onto different queues, based on their attributes.}

@@

Our algorithms for sending and receiving data implicitly perform batching and reduce the usage of system calls, and we combine this with \texttt{netmap} APIs to minimize the interaction with the OS. This allows communication performance in our framework to better scale with the number of available processors. Instead of performing the communication operations separately in the OS, by integrating them within the stream application\comment{using Click's elements within our infrastructure,}, we are able to construct a unified stream graph that represents the entire application, including computation and communication filters. Using this, we are able to parallelize these stream applications and achieve speedups of more than a factor of eight in all the applications we tested.

By integrating these features into a runtime framework, the system of multiple queues, parallel communication and computation are entirely hidden from the programmer, who merely specifies a standard stream graph for computation alone. This frees up the programmer from the onerous task of dealing with the substantial communication requirements that result from stream application distribution.

>>>>

In this paper we propose a parallel runtime framework for building stream applications that are bound by very large communication. A simple framework is used, where the filters are described as \textit{actors} and the application is represented by connecting such \textit{actors} into a directed acyclic graph (DAG). Then our framework is used to establish, (1)the communication system that allows these \textit{actors} to send and receive data across each other and (2)the computation system that executes these tasks in parallel on a given multicore system. The main contributions can be described as,

@@

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. By using the multiple queues makes it possible to the communication in parallel and by using the multiple cores, the filters from the stream application can be executed concurrently.

A major bottleneck for communication on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research uses alternate mechanisms to access data from the network hardware directly to overcome this problem\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with specialized system software that allows direct access to hardware-controlled packet queues, dramatically reducing overheads and improving performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on a \textit{flow}, which is a specific \textit{stream} between two filters. Data in a \textit{stream} must be processed sequentially, whereas different \textit{flows} can often be processed in parallel.

Newer network hardware have introduced mechanisms for separating arriving data into multiple queues, where data that belong to the same \textit{flow} are\comment{ guaranteed to be} placed in the same queue\cite{micro2008} \cite{intel2010}. We can utilize this feature to operate on \textit{flows} by processing each of the queues in parallel. However, building applications that can use this feature directly is a challenging task because it requires explicit parallel programming and interaction with low-level network interfaces. Previous research~\cite{Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207} that focuses on improving the performance of network applications in such a manner shows the complexity involved in building such a system.

@@

Our framework provides support for sending and receiving data in parallel over the NIC from user space and a common parallel execution engine that supports both computation and communication filters.

\subsection{Parallel communication interface}
\label{ppio}
The parallel communication interface, when combined with the computation filters, provides concurrent communication for the application threads or processes over the network. To do this, two important requirements have to be met. Firstly, access to network data from user space should be provided, which can be read from and written to in parallel. Secondly, it should be possible to integrate this with the computation by providing the necessary communication filters that can perform packet operations. The ability to choose communication filters allows the flexibility in the choice of communication protocol.

\subsubsection{Userspace data access}
If applications are executed in kernel space for accessing data from network devices directly would lead to scenario where the operating system can become unstable. Instead, previous researchers \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access to network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in kernel space and user space to export the network hardware buffer's memory region. This enables applications in user space to directly access the network hardware. It also prevents applications from accessing memory outside valid regions by checking the limits of the ring buffer. Packets can be written to the buffers when space is available, and an \textit{ioctl} call to the driver enables the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in user space to check when new packets are available. On reading or writing packets, the available space is updated.

@@

\subsubsection{Packet processing operations}

Packet processing operations consist of converting the raw data that the application provides into network packets suitable for transmission. Since the application tasks are represented as a set of operations on a data stream, we include these operations as stages through which the data flows through to get transformed into data packets. This is done by representing these operations as communication filters pre-built into our runtime system. Integrating them with the application task graph makes it easier to extract parallelism later in the execution engine. Several user space libraries exist that perform packet related processing operations like libnids \cite{libnids}, lwip \cite{Dunkels01designand}, Click \cite{Kohler2000} etc. Since our application is targeted at a closed environment, we chose the Click modular router due to its flexibility and vast library of elements. Our current framework supports the UDP protocol but Click's modular infrastructure makes it easy to support more complex ones. \comment{Indeed, stream applications with more complex protocols would benefit even more from the framework due to the parallelizing of these complex tasks.}

@@

In order to prevent unwanted process migration due to the OS scheduler, we specify \textit{affinity} of the MPI processes to particular CPU cores. This prevents each process from being executed on a different CPU other than that specified. This makes the processes act as execution engines, running on each of the cores on which we schedule the operations that are to be performed. Previous research \cite{Salehi:1996:EAS:234766.234769}\cite{1409136} highlights the importance of affinity in network sensitive applications. Next, by utilizing our \textit{Parallel communication interface}~(\ref{ppio}), we specify the hardware queues each of these processes handle. \comment{Initially we divide the queues evenly among the different processes we create, but later during execution we could modify the allocation based on the application operations scheduled on them. Our modifications to \texttt{netmap}'s hardware queue assignment API can help us here in redistributing the queues to different processes.} By using MPI's \texttt{send} and \texttt{receive} inter-process communication calls to handle communication between execution engines located at each CPU, it is possible to modify the attributes of the process such as CPUs assigned or number of NIC queues handled, and even the operations the processes were initially assigned.

\subsubsection{Parallel processing of computation filters}
\label{ppt}

Once we have scheduled our execution engines on the physical system, the next step is to execute the stream graph for different \textit{flows}. To extract data parallelism, we replicate the stream graph onto multiple execution engines. This ensures operations on the same \textit{flow} are performed on the same CPU, thereby using local cache. By making sure that data belonging to a specific \textit{flow} always end up in the same execution engine, we only perform operations conforming to that stream of data. \comment{We use the ability of modern NICs to classify packets onto different queues, based on their attributes.}

@@

Our algorithms for sending and receiving data implicitly perform batching and reduce the usage of system calls, and we combine this with \texttt{netmap} APIs to minimize the interaction with the OS. This allows communication performance in our framework to better scale with the number of available processors. Instead of performing the communication operations separately in the OS, by integrating them within the stream application\comment{using Click's elements within our infrastructure,}, we are able to construct a unified stream graph that represents the entire application, including computation and communication filters. Using this, we are able to parallelize the stream applications and achieve speedups of more than a factor of eight in all the applications we tested.

By integrating these features into a runtime framework, the system of multiple queues, parallel communication and computation are entirely hidden from the programmer, who merely specifies a standard stream graph for computation alone. This frees the programmer from the onerous task of dealing with the implementation of the substantial communication requirements that result from the distribution of stream applications across multiple systems.
